---
title: "Lecture 08"
output:
  html_document:
    df_print: paged
---

## Libraries

```{r}
library(tidyverse)
```


## Read source data

```{r}
# test live share
yeast.expression <- read_csv("~/Downloads/kelliher-scer-expression-data.csv")
```


## Filter the top 1600 genes in terms of their periodic rank score (see paper for details)
```{r}
yeast.1600 <-
  filter(yeast.expression, normalized_per_rank <= 1600) |>
  mutate(normalized_per_rank = NULL)
```


## Create "long" version of expression data

Note use of `names_transform` argument to insure that the time data get treated as integer values instead of characters

```{r}
yeast.long <- 
  pivot_longer(yeast.1600, !gene_ID,
               names_to = "time",
               values_to = "expression",
               names_transform = list(time = as.integer))
```

```{r}
str(yeast.long)
```


## Draw a plot showing  expression of one gene over time
```{r}
yeast.long |>
  filter(gene_ID == "NRM1") |>
  ggplot(aes(x = time, y = expression, color=gene_ID)) +
  geom_point() + 
  geom_line()

```

## Draw a plot showingexpression of two genes over time
```{r}
yeast.long |>
  filter(gene_ID %in% c("NRM1","HTB2")) |>
  ggplot(aes(x = time, y = expression, color=gene_ID)) +
  geom_point() + 
  geom_line()

```
## Problem: Magnitude of gene expression very different for these two genes

Can see this by comparing mean and std dev for expression of these two genes

```{r}
yeast.long |>
  filter(gene_ID %in% c("NRM1", "HTB2")) |>
  group_by(gene_ID) %>%
  summarize(mean.expression = mean(expression),
            std.dev.expression = sd(expression))
```

## Solution: Put these genes on a common scale by converting data to Z-scores (mean center, scale std dev to be 1)

### "manual approach"
```{r}
yeast.std <-
  yeast.long |>
  group_by(gene_ID) |>
  mutate(std_expression = (expression-mean(expression))/sd(expression),
         expression = NULL)
  
```

### Or using the built-in `scale` function 

I show both here, but generally you'd choose one or the other approach

```{r}
yeast.std <-
  yeast.long |>
  group_by(gene_ID) |>
  mutate(std_expression = scale(expression))
  
```


## Replot with scaled data


```{r}
yeast.std |>
  filter(gene_ID %in% c("NRM1","HTB2")) |>
  ggplot(aes(x = time, y = std_expression, color=gene_ID)) +
  geom_point() + 
  geom_line()

```

## Let's add one more gene to the mix

```{r}
yeast.std |>
  filter(gene_ID %in% c("NRM1","HTB2", "ACE2")) |>
  ggplot(aes(x = time, y = std_expression, color=gene_ID)) +
  geom_point() + 
  geom_line()

```

Or using a heat-plot representation

```{r}
yeast.std |>
  filter(gene_ID %in% c("NRM1","HTB2", "ACE2")) |>
  ggplot(aes(x = time, y = gene_ID, fill=std_expression)) +
  geom_tile() + 
  scale_fill_gradient2(
    low = "cyan",
    mid = "black",
    high = "yellow",
    midpoint = 0)  

```

## Create heat plot for first 100 genes in our data frame

Illustrating how unique works

```{r}
unique(yeast.std$gene_ID)[1:100]
```


Switching from `geom_tile` to `geom_raster` because geom_raster more efficient for large heat maps (but less customizable; see docs).

Also showing how to suppress the y-axis ticks and labels

```{r}
yeast.std |>
  filter(gene_ID %in% unique(yeast.std$gene_ID)[1:100]) |>
  ggplot(aes(x = time, y = gene_ID, fill=std_expression)) +
  geom_raster() + 
  scale_fill_gradient2(low = "cyan",
                       mid = "black",
                       high = "yellow",
                       midpoint = 0) + 
  theme(axis.text.y = element_blank(), axis.ticks.y = element_blank())

```

## Reordering genes by time point of maximum expression


To find the maximum expression of each gene we could use the `max` function

```{r}
yeast.std |>
  group_by(gene_ID) |>
  summarize(max.expression = max(std_expression)) 
```

The `which.max` function tells us the index at which the maximum expression occurs

```{r}
yeast.std |>
  group_by(gene_ID) |>
  summarize(max.expression = max(std_expression),
            max.index = which.max(std_expression)) 
```

We can use this information to sort gene by their time point of maximum expression. First we sort the gene names by their index of maximum expression

```{r}
genes.by.which.max <-
  yeast.std |>
  group_by(gene_ID) |>
  mutate(max.index = which.max(std_expression)) |>
  arrange(max.index) |>
  pull(gene_ID) |>
  unique()

# show the first ten genes sorted by index of max expession
head(genes.by.which.max, n = 10)
```

Then we use the `fct_relevel` function to create a new ordering of the `gene_ID` column. 

```{r}
reordered.gene_ID <- fct_relevel(yeast.std$gene_ID, genes.by.which.max) 
yeast.std$gene_ID <- reordered.gene_ID
```


Genes will no longer be shown in  alphabetical order but using the  order specified by `genes.by.which.max` vector.

The figure below shows not only the ordered genes, but illustrates a number of other tweaks including:

  * how to change the figure height and width in the code block header
  * how to set limits on a color scale
  * how to "squash" or compress data to fit in those limits (`oob` argument)
  * how to reverse a discrete axis (`scale_y_discrete(limits=rev)`).

```{r, fig.width=3, fig.height=6}
yeast.std |>
  ggplot(aes(x = time, y = gene_ID, fill=std_expression)) +
  geom_raster() + 
  scale_fill_gradient2(low = "cyan",
                       mid = "black",
                       high = "yellow",
                       midpoint = 0, 
                       limits=c(-2,2),
                       oob = scales::squish)  +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  scale_y_discrete(limits=rev)

```

